Automatic Music Generation with Deep Learning

### 2021-06-08 papers/final/jack/paper.tex

To do: figure for encoding of notes and chords to numbers

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.7\linewidth]{lstm_model.png}
  \caption{Architecture of the LSTM model.}
  \label{lstm_model}
  \Description{LSTM model}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.7\linewidth]{bidirect_lstm_model.png}
  \caption{Architecture of the bidirectional LSTM model.}
  \label{bidirect_lstm_model}
  \Description{Bidirectional LSTM model}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.5\linewidth]{attention_model.png}
  \caption{Architecture of the bidirectional LSTM model with attention mechanism.
  The Customized Attention layer is derived from Keras Layer object.}
  \label{att_lstm_model}
  \Description{Bidirectional LSTM model with attention mechanism.}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.7\linewidth]{wavenet_model.png}
  \caption{Architecture of the WaveNet-like 1-D dilated causal convolutional neural network.}
  \label{wavenet_model}
  \Description{Architecture of the 1D CNN.}
\end{figure}

@@

Similar to the evaluation of GANs, evaluating music generating models is a challenging 
task since there is a lack of metrics that are used for supervised machine learning models.

@@

\begin{itemize}
  \tightlist
  \item batch size: 32, \textbf{64}, 128
  \item learning rate: 0.01, \textbf{0.001}, 0.0001
  \item number of filters in the first Conv1D layer: 32, \textbf{64}, 128, 256
  \item number of filters in the first Conv1D layer: 32, 64, \textbf{128}, 256
  \item number of filters in the first Conv1D layer: 32, 64, 128, \textbf{256}
  \item number of filters in the first Conv1D layer: 32, 64, 128, \textbf{256}
  \item number of filters in the first Conv1D layer: 32, 64, 128, \textbf{256}
  \item dropout rate: \textbf{0.2}, 0.3, 0.4, 0.5
\end{itemize}

The models were trained for 100 epochs.

\section{Sample Output}

A collection of music generated by the LSTM models and 1-D dilated
causal convolutional neural network is available at 
\url{https://drive.google.com/drive/folders/1SlQOI_cL452PftO3KmOYO_qglE2RotCX?usp=sharing}.

@@

>>>>

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.6\linewidth]{lstm_model.png}
  \caption{Architecture of the LSTM model.}
  \label{lstm_model}
  \Description{LSTM model}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.6\linewidth]{bidirect_lstm_model.png}
  \caption{Architecture of the bidirectional LSTM model.}
  \label{bidirect_lstm_model}
  \Description{Bidirectional LSTM model}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.6\linewidth]{attention_model.png}
  \caption{Architecture of the bidirectional LSTM model with attention mechanism.
  The Customized Attention layer is derived from Keras Layer object.}
  \label{att_lstm_model}
  \Description{Bidirectional LSTM model with attention mechanism.}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.6\linewidth]{wavenet_model.png}
  \caption{Architecture of the WaveNet-style 1-D dilated causal convolutional neural network.}
  \label{wavenet_model}
  \Description{Architecture of the 1D CNN.}
\end{figure}

@@

Similar to the evaluation of GANs, evaluating music generating models is a challenging 
due to the lack of metrics that are used for supervised machine learning models.
In this paper, we evaluated the generative model by rating their outputs against our own
reference of music.

@@

\begin{itemize}
  \tightlist
  \item batch size: 32, \textbf{64}, 128
  \item learning rate: 0.01, \textbf{0.001}, 0.0001
  \item number of filters in the first Conv1D layer: \textbf{64}, 128, 256
  \item number of filters in the first Conv1D layer: 64, \textbf{128}, 256
  \item number of filters in the first Conv1D layer: 64, 128, \textbf{256}
  \item number of filters in the first Conv1D layer: 64, 128, \textbf{256}
  \item number of filters in the first Conv1D layer: 64, 128, \textbf{256}
  \item dropout rate: \textbf{0.2}, 0.3, 0.4, 0.5
\end{itemize}

The models were trained for 100 epochs. After training, the models were tested by evaluating
 their generated outputs. The winning set of hyperparameters are the combinations that result
in interestingly sounding music with a lot of variations of notes and chords. Ten samples
were generated from each trained model and were assessed qualitatively. The Bidirectional
LSTM model with attention mechamism generated the most interesting outputs with lots of
variations in the notes and chords.

\section{Sample Outputs}

A collection of music generated by the LSTM models and 1-D dilated causal convolutional 
neural network is available at 
\sloppy
\url{https://drive.google.com/drive/u/1/folders/1IO3X1OMwHKwNm26uXmez6XGLeyNLG1PZ}.

@@

Automatic music generation is an interesting problem that has recently benefited from the
development of advanced machine learning techniques. Despite being in infancy,
research has shown that deep learning generative models could produce interesting music 
with little to no human intervention. In this project, we experimented with four different
deep learning architectures, LSTM, bidirectional LSTM, bidirectional LSTM with attention
mechanism, and a WaveNet-inspired 1-D dilated causal convolutional neural network. By 
qualitative evaluation, we determined that the bidirectional LSTM model with attention 
mechanism generates the most interesting pieces of music compared to the other models.
This work, although is not directly related to my partner Alex Kyllo work on generative
models for string-quartet-based classical music, demonstrates a different perspective 
on single-track piano music generation by treating a short sequence of musical notes
and chords as a written language sentence with vocabulary and grammar.

To expand upon this project, additional work is needed to search for more sophisticated
deep learning architectures that can capture the short-term, mid-range, and long-term 
dependencies among the notes of a musical composition. Furthermore, these architectures 
should be generalized to music that is performed by multiple instruments (multi-track 
polyphonic music)). Another improvement is to enable the generative models to work with
multiple musical genres, not just classical music. Finally, we only relied on our own
qualitative evaluation to assess the quality of the models due to the time constraint
of the project. We also need a comprehensive evaluation plan, which involves qualitative evaluation 
by the multiple human judges and quantitative evaluation on the structure of the generative 
musical sequences.
